Multilevel spin models in cold atomic systems

### 2021-04-02 sun_phases/spin_model_paper/main.tex
update SU(n) spin model paper to reflect changes to figures

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, we can prepare an $N$-fold product of the state
\begin{align}
  \ket{\x\x} \equiv e^{-\i\frac{\pi}{2}s_\y^2} \ket{s}
  \simeq \ket{+\x} + \ket{-\x},
\end{align}
where $\simeq$ denotes equality up to a global phase and normalization factor, and
\begin{align}
  \ket{\pm\x} \equiv e^{\mp\i\frac{\pi}{2}s_\y} \ket{s}
  = \f1{2^s} \sum_\mu \p{\pm1}^{s-\mu}
  { 2s \choose s+\mu }^{1/2} \ket{\mu}
\end{align}
are states of a single spin polarized in opposite directions along the $x$ axis.
Here ${m \choose k}$ is a binomial coefficient.
The state $\ket{\x\x}$ has a vanishing mean spin vector, $\bk{S_\x}=\bk{S_\y}=\bk{S_\z}=0$, (where $S_\alpha = \sum_j s_{\alpha j}$), but variances $\bk{S_\x^2}=N\times s^2$ and $\bk{S_\y^2}=\bk{S_\z^2}=N\times s/2$.

@@

\begin{figure}
\centering
\includegraphics{mean_op_ss_X.pdf}
\caption{
Long-time averages of $\bar\sigma_\x$ and $\ss$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket{+\x}^{\otimes N}$ up to a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $\bbk{\bar\sigma_\x}_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_op_ss_X}
\end{figure}

Figure \ref{fig:mean_op_ss_X} shows the long-time averages of the transverse magnetization $\bar\sigma_\x$ and interaction energy $\ss$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket{+\x}^{\otimes N}$.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero transverse magnetization $\bbk{\sigma_\x}_\MF$ and a large interaction energy $\bbk{\ss}_\MF$, while the disordered phase has no transverse magnetization and an interaction energy close to its minimal value.
Numerically, we find that the transition between these phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields_X}).

\begin{figure}
\centering
\includegraphics{crit_fields_X.pdf}
\caption{
The critical value of $\p{J\phi/U}_{\t{crit}}$ as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket\x^{\otimes N}$.
A single-parameter fit to $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ finds $\alpha=0.333(5)$, and $\alpha=1/3$ is consistent with all mean-field results to within an uncertainty determined by the resolution of $J\phi/U$ in mean-field simulations.
}
\label{fig:crit_fields_X}
\end{figure}

As shown in insets of Figure \ref{fig:mean_op_ss_X}, mean-field results for different spin dimensions $n$ collapse onto each other when normalizing the field $J\phi/U$ to its critical value, $\p{J\phi/U}\to\p{J\phi/U}\times\p{n/2}^{1/3}$, and rescaling
\begin{align}
  \bbk{\bar\sigma_x}_\MF \to \f{\bbk{\bar\sigma_x}_\MF}{\gamma\p{n/2}},
  &&
  \bbk{\ss}_\MF \to \f{\bbk{\ss}_\MF-\gamma\p{n}}{1-\gamma\p{n}},
  \label{eq:rescale}
\end{align}
where
\begin{align}
  \gamma\p{k} \equiv \f{\Gamma\p{k-\frac12}}{\sqrt\pi\,\Gamma\p{k}}
  \stackrel{k\ge2}{\approx} \f1{\sqrt{\pi(k-1)}}.
  \label{eq:gamma}
\end{align}
The rescaling of transverse magnetization and interaction energy can be understood by considering their limiting behavior as $J\phi/U\to\infty$ or $J\phi/U\to0^+$.

In the strong-field limit $J\phi/U\to\infty$, we can ignore interactions and treat spins as though they simply precess at different rates.
The long-time-averaged transverse magnetization $\bbk{\bar\sigma_\x}_\MF$ then trivially vanishes as $J\phi/U\to\infty$.
The interaction energy $\bk{\ss}_\MF = \bk{\bar{\v s}}_\MF \c \bk{\bar{\v s}}_\MF$, meanwhile, has contributions from
\begin{enumerate*}
\item the diagonal parts of the mean spin matrix $\bk{\bar{\v s}}_\MF$, which are conserved by inhomogeneous spin precession, and
\item the off-diagonal parts of $\bk{\bar{\v s}}_\MF$, whose oscillations average to zero when evaluating the long-time average $\bbk{\ss}_\MF$.
\end{enumerate*}
Altogether, the interaction energy $\bbk{\ss}_\MF$ in the strong-field limit is determined by the time-independent diagonal part $\diag\bk{\bar{\v s}\p{t}} = \diag\op\x$, namely
\begin{align}
  \lim_{J\phi/U\to\infty} \bbk{\ss}_\MF
  = \Tr\sp{\p{\diag\op\x}^2}
  = \gamma\p{n}.
\end{align}
The same result can be obtained by computing the long-time-averaged interaction energy of two spins precessing at different rates.

In the weak-field limit $J\phi/U\to0^+$, the spin-locking $\v S\c\v S$ interactions of the spin Hamiltonian $H_{\t{spin}}$ effectively restrict dynamics to the permutationally-symmetric (PS) manifold.
At second order in perturbation theory, the effective spin Hamiltonian within the PS manifold becomes (see Appendix \ref{sec:pert_theory})
\begin{align}
  H_{\t{spin}}^{\t{eff}}
  = \f{\p{J\phi}^2}{\p{N-1}U}
  \times \sp{S_\z^2 - N\sum_q s_{\z,q}^2},
\end{align}
whose mean-field version is
\begin{align}
  H_{\t{MFT}}^{\t{eff}} = -\f{\p{J\phi}^2}{U} \sum_q s_{\z,q}^2,
\end{align}
where we have used the fact that the axial magnetizations $\bk{s_{\z,q}}=0$ are conserved.
The weak-field effective Hamiltonian thus preserves permutational symmetry, so $\bbk{\ss}_\MF\to1$ as $J\phi/U\to0^+$, while the transverse magnetization $\bbk{\bar\sigma_\x}_\MF$ approaches the long-time average of $\sigma_\x\equiv s_\x/s$ for a single spin:
\begin{align}
  \lim_{J\phi/U\to0^+} \bbk{\bar\sigma_\z}_\MF
  = \lim_{T\to\infty} \f1T \int_0^T \dd\tau
  \bk{\x|\sigma_\x\p{\tau}|\x},
\end{align}
where
\begin{align}
  \sigma_\x\p{\tau} = e^{\i\tau s_\z^2} \sigma_\x e^{-\i\tau s_\z^2}.
\end{align}
By mapping the Hilbert space of an $n$-level spin to the Dicke (PS) manifold of $n-1$ spin-$\frac12$ particles, we can adapt exact analytical results for the Ising model \cite{foss-feig2013nonequilibrium}\footnote{See Appendix K of Ref.~\cite{perlin2020shorttime} for a simpler adaptation of the analytics in Ref.~\cite{foss-feig2013nonequilibrium} to the one-axis twisting model $H_{\t{OAT}}=\chi s_\z^2$.} to find that
\begin{align}
  \bk{\x|\sigma_\x\p{\tau}|\x} = \p{\cos\tau}^{n-2},
\end{align}
so for even $n$
\begin{align}
  \lim_{J\phi/U\to0^+} \bbk{\bar\sigma_\z}_\MF
  = \f1{2\pi} \int_0^{2\pi} \dd\tau \p{\cos\tau}^{n-2}
  = \gamma\p{\f{n}{2}}.
\end{align}

>>>>

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ prior to evolution under the ``gauged frame'' spin Hamiltonian $H_{\t{spin}}$ in Eq.~\eqref{eq:H_spin} or the three-laser drive $H_{\t{3LD}}$ in Eq.~\eqref{eq:drive_all}.
We assume the capability to prepare an initial state in which all spins are maximally polarized along $\v v_+$, i.e.~$\ket{\v v_+}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauged frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, we can prepare an $N$-fold product of the state
\begin{align}
  \ket{\x\x} \equiv e^{-\i\frac{\pi}{2}s_\y^2} \ket{s}
  \simeq \ket{+\x} + \ket{-\x},
\end{align}
where $\simeq$ denotes equality up to a global phase and normalization factor, and
\begin{align}
  \ket{\pm\x} \equiv e^{\mp\i\frac{\pi}{2}s_\y} \ket{s}
  = \f1{2^s} \sum_\mu \p{\pm1}^{s-\mu}
  { 2s \choose s+\mu }^{1/2} \ket{\mu}
\end{align}
are states of a single spin polarized in opposite directions along the $x$ axis.
Here ${m \choose k}$ is a binomial coefficient.
The state $\ket{\x\x}$ has a vanishing mean spin vector, $\bk{S_\x}=\bk{S_\y}=\bk{S_\z}=0$ (where $S_\alpha = \sum_j s_{\alpha j}$), but variances $\bk{S_\x^2}=N\times s^2$ and $\bk{S_\y^2}=\bk{S_\z^2}=N\times s/2$.

@@

\begin{figure}
\centering
\includegraphics{mean_sx_ss.pdf}
\caption{
Long-time averages of $\bar\sigma_\x$ and $\ss$ for different spin dimensions $n$ (indicated in the legend) as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket{+\x}^{\otimes N}$ up to a time $T=10^5/U$.
Insets show same data after normalizing $J\phi/U\to J\phi/U \times (n/2)^{1/3}$, as well as transforming $\bbk{\bar\sigma_\x}_\MF$ and $\bbk{\ss}_\MF$ according to Eq.~\eqref{eq:rescale}.
}
\label{fig:mean_sx_ss}
\end{figure}

Figure \ref{fig:mean_sx_ss} shows the long-time averages of the transverse magnetization $\bar\sigma_\x$ and interaction energy $\ss$ as computed by mean-field simulations of $N=100$ spins in an initial $x$-polarized state $\ket{+\x}^{\otimes N}$.
As expected, the spin model exhibits a mean-field dynamical phase transition between an ordered phase at small $J\phi/U$ and a disordered phase at large $J\phi/U$.
The ordered phase has a non-zero transverse magnetization $\bbk{\sigma_\x}_\MF$ and an interaction energy $\bbk{\ss}_\MF$ that asymptotically approaches its maximal value as $J\phi/U\to0^+$.
The disordered phase has no transverse magnetization, $\bbk{\bar\sigma_\x}_\MF=0$, but an interaction energy $\bbk{\ss}_\MF$ that nonetheless indicates persistent (albeit small) nontrivial inter-spin correlations when $n>2$.
These nontrivial inter-spin correlations vanish as $J\phi/U\to\infty$, in which case $\bbk{\ss}_\MF$ approaches the minimal value allowed by conservation laws (clarified below).
By minimizing the reduced field $J\phi/U$ for which $\bbk{\sigma_\x}_\MF=0$, we numerically find that the transition between ordered and disordered phases occurs at a critical field $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ with $\alpha\approx1/3$ (see Figure \ref{fig:crit_fields}).

\begin{figure}
\centering
\includegraphics{crit_fields.pdf}
\caption{
The critical value of $\p{J\phi/U}_{\t{crit}}$ as determined by mean-field simulations of $N=100$ spins initially in the $x$-polarized state $\ket\x^{\otimes N}$.
A single-parameter fit to $\p{J\phi/U}_{\t{crit}}=\p{n/2}^{-\alpha}$ finds $\alpha=0.333(5)$, and $\alpha=1/3$ is consistent with all mean-field results to within an uncertainty determined by the resolution of $J\phi/U$ in mean-field simulations.
}
\label{fig:crit_fields}
\end{figure}

As shown in insets of Figure \ref{fig:mean_sx_ss}, mean-field results for different spin dimensions $n$ collapse onto each other when normalizing the field $J\phi/U$ to its critical value, $\p{J\phi/U}\to\p{J\phi/U}\times\p{n/2}^{1/3}$, and rescaling
\begin{align}
  \bbk{\bar\sigma_x}_\MF \to \f{\bbk{\bar\sigma_x}_\MF}{\gamma\p{n/2}},
  &&
  \bbk{\ss}_\MF \to \f{\bbk{\ss}_\MF-\gamma\p{n}}{1-\gamma\p{n}},
  \label{eq:rescale}
\end{align}
where
\begin{align}
  \gamma\p{k} \equiv \f{\Gamma\p{k-\frac12}}{\sqrt\pi\,\Gamma\p{k}}
  \stackrel{k\ge2}{\approx} \f1{\sqrt{\pi(k-1)}}.
  \label{eq:gamma}
\end{align}
The rescaling of transverse magnetization and interaction energy can be understood by considering their limiting behavior as $J\phi/U\to\infty$ or $J\phi/U\to0^+$.

In the strong-field limit $J\phi/U\to\infty$, we can ignore interactions and treat spins as though they simply precess at different rates.
The long-time-averaged transverse magnetization $\bbk{\bar\sigma_\x}_\MF$ then trivially vanishes as $J\phi/U\to\infty$.
The interaction energy $\bk{\ss}_\MF = \bk{\bar{\v s}}_\MF \c \bk{\bar{\v s}}_\MF$, meanwhile, has contributions from
\begin{enumerate*}
\item the diagonal parts of the mean spin matrix $\bk{\bar{\v s}}_\MF$, which are conserved by inhomogeneous spin precession, and
\item the off-diagonal parts of $\bk{\bar{\v s}}_\MF$, whose oscillations average to zero when evaluating the long-time average $\bbk{\ss}_\MF$.
\end{enumerate*}
Altogether, the interaction energy $\bbk{\ss}_\MF$ in the strong-field limit is determined by the time-independent diagonal part $\diag\bk{\bar{\v s}\p{t}} = \diag\op{+\x}$, namely
\begin{align}
  \lim_{J\phi/U\to\infty} \bbk{\ss}_\MF
  = \Tr\sp{\p{\diag\op{+\x}}^2}
  = \gamma\p{n}.
\end{align}
The same result can be obtained by computing the long-time-averaged interaction energy of two spins precessing at different rates.

In the weak-field limit $J\phi/U\to0^+$, the spin-locking $\v S\c\v S$ interactions of the spin Hamiltonian $H_{\t{spin}}$ effectively restrict dynamics to the permutationally-symmetric (PS) manifold.
At second order in perturbation theory, the effective spin Hamiltonian within the PS manifold becomes (see Appendix \ref{sec:pert_theory})
\begin{align}
  H_{\t{spin}}^{\t{eff}}
  = \f{\p{J\phi}^2}{\p{N-1}U}
  \times \sp{S_\z^2 - N\sum_q s_{\z,q}^2},
\end{align}
which at the mean-field level becomes
\begin{align}
  H_{\t{MFT}}^{\t{eff}} = -\f{\p{J\phi}^2}{U} \sum_q s_{\z,q}^2,
\end{align}
where we have used the fact that the axial magnetizations $\bk{s_{\z,q}}=0$ are conserved.
The weak-field effective Hamiltonian thus preserves permutational symmetry, so $\bbk{\ss}_\MF\to1$ as $J\phi/U\to0^+$, while the transverse magnetization $\bbk{\bar\sigma_\x}_\MF$ approaches the long-time average of $\sigma_\x\equiv s_\x/s$ for a single spin:
\begin{align}
  \lim_{J\phi/U\to0^+} \bbk{\bar\sigma_\z}_\MF
  = \lim_{T\to\infty} \f1T \int_0^T \dd\tau
  \bk{+\x|\sigma_\x\p{\tau}|+\x},
\end{align}
where
\begin{align}
  \sigma_\x\p{\tau} = e^{\i\tau s_\z^2} \sigma_\x e^{-\i\tau s_\z^2}.
\end{align}
By mapping the Hilbert space of an $n$-level spin onto the PS (Dicke) manifold of $n-1$ spin-$\frac12$ particles, we can adapt exact analytical results for the Ising model \cite{foss-feig2013nonequilibrium}\footnote{See Appendix K of Ref.~\cite{perlin2020shorttime} for a simpler adaptation of the analytics in Ref.~\cite{foss-feig2013nonequilibrium} to the one-axis twisting model $H_{\t{OAT}}=\chi s_\z^2$.} to find that
\begin{align}
  \bk{+\x|\sigma_\x\p{\tau}|+\x} = \p{\cos\tau}^{n-2},
\end{align}
so for even $n$
\begin{align}
  \lim_{J\phi/U\to0^+} \bbk{\bar\sigma_\z}_\MF
  = \f1{2\pi} \int_0^{2\pi} \dd\tau \p{\cos\tau}^{n-2}
  = \gamma\p{\f{n}{2}}.
\end{align}
